A new tidy data structure to support exploration and modeling of temporal data

### 2019-01-18 main.Rmd
section 2.1 done

The data problems are grouped into two types of analysis, time series and longitudinal. Despite of exactly the same data input, the representation of time series and longitudinal data diverges due to different modelling approaches.

Time series can be univariate or multivariate, and for modelling requires relatively large in length (i.e. large $T$). Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic processes, for the primary purpose of forecasting, and characterizing temporal dynamics. The time series supporting modeling are represented as vectors or matrices in most statistical software. Multivariate time series are typically assumed to be in the format where each row is assumed to hold observations at a time point and each column to contain a single time series. (The tidy data name for this would be **wide format**.) This implies that data are columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric variables. In addition, time indexes are stripped off the data and implicitly inferred as attributes or meta-information. It strictly requires that the number of observations must be the same across all the series. Data wrangling from the form that data arrives in, to this specialist format can be frustrating and difficult, inhibiting the variety of downstream tasks such as analytics.

For longitudinal analysis, researchers and analysts are primarily interested in explaining trends across and variations among individuals, and making inference about a broader population. Longitudinal data or panel data typically assumes fewer measurements (small $T$) over a large number of individuals (large $N$). It often occurs that measurements for individuals are taken at different time points, resulting in an unbalanced panel. Thus, the primary format required for modeling is stacked series, blocks of measurements for each individual, with columns indicating individual, time of measurement and the measurements. (The tidy data name for this would be **long format**.) Evidently, this data organisation saves storage space for many sparse cells, compared to structuring it in that wide format which would have missing values in many cells. A detriment of this format is that demographic information for subjects is often repeated for each time point. However, appealing feature is that data is structured in a concise and semantic manner with reference to observations and variables, with the time index stated explicitly. This opens the door to easily operating on time to make calculations and extract different temporal components, such as month and day of the week. It is conducive to examining the data in many different ways and leading to more comprehensive exploration and forecasting.

>>>>

The data problems are grouped into two types of analysis, time series and longitudinal. Despite being exactly the same data input, the representation of time series and longitudinal data diverges due to different modelling approaches.

Time series can be univariate or multivariate, and for modelling requires relatively large in length (i.e. large $T$). Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic processes, for the primary purpose of forecasting, and characterizing temporal dynamics. The time series supporting modeling are represented as vectors or matrices in most statistical software. Multivariate time series are typically assumed to be in the format where each row is assumed to hold observations at a time point and each column to contain a single time series. (The tidy data name for this would be **wide format**.) This implies that data are columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric variables. In addition, time indexes are stripped off the data and implicitly inferred as attributes or meta-information. It strictly requires that the number of observations must be the same across all the series. Data wrangling, from the form that data arrives in, to this specialist format, can be frustrating and difficult, inhibiting the variety of downstream tasks such as analytics.

For longitudinal analysis, researchers and analysts are primarily interested in explaining trends across and variations among individuals, and making inference about a broader population. Longitudinal data or panel data typically assumes fewer measurements (small $T$) over a large number of individuals (large $N$). It often occurs that measurements for individuals are taken at different time points, resulting in an unbalanced panel. Thus, the primary format required for modeling is stacked series, blocks of measurements for each individual, with columns indicating individual, time of measurement and the measurements. (The tidy data name for this would be **long format**.) Evidently, this data organisation saves storage space for many sparse cells, compared to structuring it into wide format which would have missing values in many cells. A drawback of this format is information unique to each individual is often repeated for all time points. The appealing feature is that data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. This opens the door to easily operating on time to make calculations and extract different temporal components, such as month and day of the week. It is conducive to examining the data in many different ways and leading to more comprehensive exploration and forecasting.
